### 2022-06-23 manuscript/hiv-gwas.tex
Update on Overleaf.

% \begin{flushright} % for MBE submission
% Submission type: Article\\
% Section: Methods
% \end{flushright}

@@

% Overview of relevant methods and why we don't use them
It is well-known that GWAS can be sensitive to confounding variables. Shared ancestry among individuals, especially between close relatives, can give rise to spurious genetic correlations with a trait. Corrections for these types of population structure in human GWAS cohorts are well-developed and widely accepted  \citep{Astle2009, Price2006}. More recently, analogous methods have been developed for microbial GWAS, where clonal reproduction exacerbates population structure \citep{Power2017}. Microbial GWAS-specific phylogenetic methods to account for population structure in microbial GWAS include explicitly testing for lineage-specific effects as in \citep{Earle2016} and modified association tests that account for phylogenetic relationships amongst samples as in \citep{collins_phylogenetic_nodate}. 
% \citep{Earle2016} simultaneously test for pathogen lineage and locus-level associations with a trait using a linear mixed model framework. \citep{collins_phylogenetic_nodate} test for locus-level associations that are significantly stronger than expected under a null model considering shared pathogen ancestry. 
These approaches are designed to quantify genetic effects from one organism on a trait.

In the infectious disease context, genetic effects from two organisms - the host and the pathogen - may affect an infectious disease trait. GWAS using paired host-pathogen genotype data have previously been done to elucidate the marginal and interaction effects of host and pathogen genetic variants. Methods to account for microbial population structure when testing for marginal host associations or host-pathogen interaction effects include adding the microbial kinship matrix as a random effect in a linear mixed model as in \citep{Wang2018Two-wayGenomes} and using principle components derived from either this matrix or the pathogen phylogeny as covariates in a linear model as in \citep{Naret2018}. These methods focus on capturing and accounting for correlations due to the pathogen phylogeny, without further investigating the nature of these correlations.
%However, the top principle components capture only some of the information from the full pathogen phylogeny. Furthermore, we would like to simultaneously address issues (i) and (ii).

@@

\begin{figure}[H]
	\begin{center}
	\includegraphics[width = \linewidth]{figures/simulation_setup.png}
	\caption{A high-level schematic of our POUMM-based simulation framework in the context of HIV-1 spVL. (A) shows how the viral effects on spVL evolve along the viral phylogeny according to an Ornstein-Uhlenbeck process. (B) shows how human host genetic effects are the sum of independent effects from several causal variants. Each variant can be present in 0, 1, or 2 copies. Half the variants have a positive effect of size $\delta$ and half have a negative effect of size $\delta$. (C) shows how other environmental effects are independently drawn from a Gaussian distribution centered at 0. These three effects sum to the trait value for each simulated individual.}
		\label{fig:spVLModel}
	\end{center}
\end{figure}

@@

\begin{enumerate}
	\item Sample paired host genotypes, pathogen genome sequences, and trait values from a cohort.
	\item Construct a pathogen phylogeny using the pathogen genome sequences.
	\item Estimate the parameters of the POUMM based on the trait values and the pathogen phylogeny. This can be done with e.g. the R package POUMM \citep{Mitov2017a-POUMM}.
	\item Generate maximum-likelihood estimates for the pathogen and corresponding non-pathogen effects on the trait using equations \ref{eq:MuG} and \ref{eq:EHat}.
	\item Perform GWAS with only the non-pathogen effects on the trait as the response variable.
\end{enumerate}

@@

\begin{figure}[H]
	\begin{center}
		\includegraphics[width=\linewidth]{figures/simulation_results}
		\caption{Results from the simulation study. We simulated host, pathogen, and environmental effects on a trait under the POUMM with different heritability (y-axis) and selection strength (x-axis) parameters. For each simulated dataset, we applied our method to estimate the non-pathogen effects and performed GWAS with these values. (A) shows the root mean squared error (RMSE) of our estimator (left) compared to un-corrected trait values, scaled by their mean (right) under each simulated evolutionary scenario. The RMSE is with reference to the true (simulated) host part of the trait values. Thus, more accurate estimates (lower RMSE) mean the trait value used for GWAS will be closer to the true host part of the trait value. (B) shows how GWAS power can improve given the true, simulated non-pathogen effect on spVL (left) and using our estimate for this value (middle) compared to using the scaled trait value (right). Each tile's color corresponds to the average value across 20 simulated datasets of 500 samples. The points highlight specific heritability and selection strength values from the \emph{A. thaliana}-\emph{X. arboricola} QDR analysis, HIV-1 spVL analysis, and four simulated scenarios that are presented in more detail in Figure \ref{fig:tip-trait-correlation}. RMSE = root mean square error, TPR = true positive rate, QDR = quantitative disease resistance, spVL = set-point viral load.}
		\label{fig:simulationResults}
	\end{center}
\end{figure}

\begin{figure}[H]
	\begin{center}
		\includegraphics[width=\linewidth]{figures/example_simulation_trait_values}
		\caption{Simulated data from two evolutionary scenarios where a phylogenetic correction to trait values improves GWAS power (right side) and where it does not (left side). These examples correspond to two of the unfilled points in Figure \ref{fig:simulationResults}. (A) and (B) show total trait values for 12 randomly selected tips from the simulated phylogeny with pathogen heritability 15 and 75\%, respectively. Depending on the pathogen heritability, trait values are more or less correlated at clustered tips. (C) compares our method's estimate for the non-pathogen part of trait values (y-axis) with true simulated host trait values (x-axis) with pathogen heritability 15 and 75\%. The solid line is the y=x line. Selection strength was fixed to 0.1 time $^{-1}$ for both scenarios and all other parameters were fixed as in the full simulation study.}
		\label{fig:simulationTraitExample}
	\end{center}
\end{figure}

\begin{figure}[H]
\begin{center}
\includegraphics[width = \linewidth]{figures/tip_trait_correlation.png}
	\caption{Correlations between trait values in pairs of tips in four simulated scenarios. These examples correspond to the four unfilled points in Figure \ref{fig:simulationResults}. Correlations are calculated for pairs of tips binned by phylogenetic distance (into deciles) across the 20 replicate simulations for each of the four evolutionary scenarios. Trait values are only noticeably correlated for closely clustered tips under the scenario with high pathogen heritability and low selection strength/ low stochastic fluctuations (upper left facet).}
	\label{fig:tip-trait-correlation}
	\end{center}
\end{figure}

@@

\begin{figure}[H]
\begin{center}
\includegraphics[width = \linewidth]{figures/application_experimental_setups.pdf}
	\caption{A high-level schematic of the experimental setup for the two application datasets. For (A) HIV-1 spVL in the Swiss HIV Cohort Study, data are paired viral and human genotypes and associated spVL measurements. We fit the POUMM to the viral phylogeny and spVL values associated with each infected individual ($z_1$, $z_2$, ..., $z_{1493}$). For (B) \emph{A. thaliana}-\emph{X. arboricola} QDR from \cite{Wang2018Two-wayGenomes}, data are bacterial and plant genotypes with QDR measurements for all possible combinations of pathogen and host plant strains. We fit the POUMM to the bacterial phylogeny and mean QDR calculated for each pathogen strain across all the hosts plant types ($\bar{z}_1$, $\bar{z}_2$, ..., $\bar{z}_{22}$).}
	\label{fig:application-exp-setup}
	\end{center}
\end{figure}

We compared our proposed GWAS framework with a more standard approach by performing two different GWAS on the same SHCS human genotypes. We retained 1,392 individuals of European ancestry for the GWAS. In the (i) ``GWAS with standard trait value'' we used the total trait value, our calculated spVL values, as the GWAS response variable. In the (ii) ``GWAS with estimated non-pathogen part of trait'' we used our estimates for the non-pathogen effects on spVL. Figure \ref{fig:gwas-results}A shows that results are qualitatively similar between the two GWAS. Q-Q plots show the distribution of p-values are very similar as well (Figure \ref{fig:qq-plots}). Figure \ref{fig:gwas-results}B shows how the strength of association changed for some variants in the MHC and \emph{CCR5} regions. Taking into account phylogenetic information slightly decreased association strength for most variants in the \emph{CCR5} region. Association strength increased for some variants in the MHC, for example, SNP rs9265880 had the greatest increase in significance in the MHC region, from a p-value of $3.5 \times 10^{-07}$ to $7.7 \times 10^{-09}$. However, the top-associated variants in the MHC and \emph{CCR5} regions were consistent regardless of the GWAS response variable used (Table \ref{tab:gwas-top-associations}). Finally, Table \ref{tab:comp-gwas-mclaren-snps} shows how our GWAS results compare for the two top-associated SNPs identified by \citet{McLaren2015}, who performed the largest standard GWAS for HIV spVL to date. We repeated the analysis using three different approximate maximum-likelihood phylogenies and these results were consistent (see Materials and Methods; Table \ref{tab:comp-gwas-mclaren-snps-alt-trees}). In summary, there are no clear patterns that point to new regions of association in the human genome with spVL when we take into account the pathogen phylogeny.

\begin{figure}[H]
	\begin{center}
		\includegraphics[width=\linewidth]{figures/gwas_results}
		\caption{Results from comparative GWAS on HIV-1 spVL data. (A) shows association p-values for the same host variants from the SHCS cohort in GWAS with two different response variables. On the left, we used unmodified (total) spVL values. On the right, we used our estimates for the non-pathogen effects on spVL. The alternating shades correspond to different chromosomes. (B) compares the strength of association for variants in the \emph{CCR5} and MHC regions between the two GWAS (positions 45.4 - 47Mb on chromosome 3 and 29.5 - 33.5Mb on chromosome 6 for the \emph{CCR5} and MHC, respectively). Base positions are with reference to genome build GRCh37. The color of each point represents the difference in -log$_{10}$ p-value between the two GWAS. Red means taking into account phylogenetic information decreased the strength of association and blue means it increased it. The dashed lines show genome-wide significance at p = $5 \times 10^{-8}$.}
		\label{fig:gwas-results}
	\end{center}
\end{figure}

@@

Given the posterior mean estimates for the POUMM parameters, expected correlation in trait values between tips were very low (maximum value $3.2 \times 10^{-12}$ compared to maximum value of 0.45 in the HIV-1 spVL application). Thus, the phylogeny is not very informative for a trait value correction. Indeed, the estimated pathogen part of the QDR trait calculated by our method is simply a scaling of the total QDR trait value (Figure \ref{fig:trait-comparison-qdr}). We anyways selected 22 random host-pathogen strain pairings to perform a comparative GWAS analogous to that for HIV-1 spVL, where each host is infected with a single pathogen strain. In the first GWAS, we used the specific QDR measurement for each selected host-pathogen pairing. I.e., with reference to Figure \ref{fig:application-exp-setup}, we selected $z_{11}$ for the first sample, $z_{23}$ for the second sample, and so on. In the second GWAS, we used our estimates for the non-pathogen effects on QDR for each pairing. Since our method did not utilize phylogenetic information in this case, the estimated non-pathogen part of the trait is simply the specific QDR for each selected host-pathogen pairing, minus mean QDR for the respective pathogen strain, calculated across all the host \emph{A. thaliana} types. I.e., with reference to Figure \ref{fig:application-exp-setup}, we used a scaled version of $z_{11} - \bar{z}_1$ for the first sample, $z_{23} - \bar{z}_2$ for the first sample, and so on. Figure \ref{fig:gwas-results-qdr} shows that results are qualitatively similar between the two GWAS, with a slight decrease in association strength for the top-associated variants. Q-Q plots show the distribution of p-values are also very similar (\ref{fig:qq-plots-qdr}). In the first, standard GWAS, one \emph{A. thaliana} loci just exceeds the threshold for significant association after correction for multiple testing. In the second, corrected GWAS, no \emph{A. thaliana} variants are significantly associated with QDR to \emph{X. arboricola}.

\begin{figure}[H]
\begin{center}
\includegraphics[width = \linewidth]{figures/gwas_results_xanthomonas_maxmaf.png}
	\caption{Results from comparative GWAS on \emph{A. thaliana}-\emph{X. arboricola} QDR data. The two facets show association p-values for the same host \emph{A. thaliana} variants in GWAS with two different response variables. On the left, we used unmodified (total) QDR values for each of the 22 selected host-pathogen pairings on which these results are based. On the right, we used our estimates for the non-pathogen effects on QDR for these samples. In this case, estimated non-pathogen effects are the specific QDR for each selected host-pathogen pairing, minus mean QDR for the respective pathogen strain, calculated across all the host \emph{A. thaliana} types. The alternating shades correspond to different chromosomes. The dashed lines show significance at significance level 0.05 with a Bonferroni correction for multiple testing.}
	\label{fig:gwas-results-qdr}
	\end{center}
\end{figure}

@@

% Contextualize the study & address limitations
The main novelty of our approach is to estimate parameters governing the evolutionary dynamics of a trait in the pathogen population and use these estimates to correct infectious disease trait values prior to performing GWAS, thereby estimating and removing pathogen effects. In simulations, we show that when trait heritability due to shared pathogen ancestry amongst infection partners is greater than approximately 25\%, GWAS power to detect host genetic variants associated with the same trait is reduced. Our method can correct for this effect in certain evolutionary scenarios by using information from the full pathogen phylogeny. Based on our simulation results, our method is anticipated to be very useful for disease traits that are highly heritable from donor to recipient and  maintain a high correlation between sampled individuals. In simulations, we showed this is the case when pathogen heritability is high, selection strength is low, and trait values are not subject to strong stochastic fluctuations. It is possible that traits under strong selection pressure are also highly heritable and phylogenetically correlated, for instance antimicrobial resistance. However, such traits might violate the POUMM assumption that trait values vary as a random walk in continuous space if they are caused by few mutations of strong affect. In summary, cohort-level, phylogenetically structured differences in the measured trait value are necessary for our approach to outperform state of the art methods.

@@

Given our estimates for trait heritability and selection strength on HIV-1 spVL and \emph{A. thaliana} quantitative disease resistance to \emph{X. arboricola}, our simulation results reveal that we cannot expect a significant improvement in GWAS power for these systems (Figure \ref{fig:simulationResults}). Indeed, while certain pairs of samples in the HIV-1 cohort were expected to have phylogenetically correlated spvL values (maximum expected correlation between any two samples was 0.45), the overall effect on GWAS is small. For HIV-1 spVL, our phylogenetic correction slightly decreases p-values for variants in $CCR5$ and slightly decreases some and increases other p-values for variants in the MHC (Figure \ref{fig:gwas-results}B). Simulations show we shouldn't expect a net p-value decrease, but our simulations represent an ideal scenario since we simulate under the POUMM. For the empirical data, un-modeled evolutionary pressures like drug treatment and host-specific HLA alleles might cause the reduced p-values. However, the overall picture is consistent between the two GWAS (Figure \ref{fig:gwas-results}A). For \emph{A. thaliana} QDR to \emph{X. arboricola}, the trait value correction does not utilize phylogenetic information because phylogenetic correlations between samples are too weak (maximum expected correlation between strains was $3.2 \times 10^{-12}$). We anyways corrected QDR trait values based on average QDR for each pathogen strain across the full range of host types. Results show slight decrease in p-values for the most-associated variants in this application as well, but the overall picture is consistent with previous GWAS results from \cite{Wang2018Two-wayGenomes}. That study found no significant \emph{A. thaliana} variants associated with QDR using a linear mixed model jointly accounting for host genetic effects, pathogen genetic effects, and interaction effects. As with HIV-1 spVL, our results do not challenge this previous finding. Therefore, we conclude that GWAS for host determinants of HIV-1 subtype B spVL and \emph{A. thaliana} determinants of QDR to \emph{X. arboricola} are robust to our correction for pathogen effects.

Our approach is envisioned to be complementary to previously developed methods for infectious disease GWAS. First, it provides additional information on the evolutionary dynamics of the trait in the pathogen population. Then, it is a convenient pre-processing step for GWAS because it simply produces a corrected response variable for GWAS association tests. As in the \emph{A. thaliana}-\emph{X. arboricola} application, fitting the POUMM may reveal that expected phylogenetic correlations between samples are not strong enough to justify using our method to correct trait values in a GWAS. In this case, one may wish to use a linear mixed model as in \citep{Wang2018Two-wayGenomes}, where the pathogen effect is co-estimated as a random effect. In cases where a correction can be estimated and applied, the corrected trait values are envisioned to be used in any of the previously developed GWAS models for the actual association testing (we used a linear model approach implemented in PLINK \citep{Chang2015}, though a more advanced method would be to use a linear mixed model with host ancestry as a random effect). Further, additional model complexity can be added to the GWAS association tests. For instance, our method does not account for co-infection, which might add additional variance to trait values and decrease GWAS power. In this case, one could add co-infection status as a covariate in the GWAS association test to account for this variable.

@@

% Summarize the paper
In summary, we present a coherent infectious disease GWAS framework that takes the pathogen phylogeny into account when searching for host determinants of a disease trait. We further show that the pathogen phylogeny only has an impact on the GWAS outputs if heritability of the trait amongst infection partners is $>$ 25\%.
This means that for the systems studied here, spVL in individuals living with HIV and QDR for X. arboricola infections in A. thaliana, the phylogenetic correction does not change GWAS results. Our findings indicate previously published GWAS results for these systems are not biased by shared evolutionary history amongst infecting pathogen strains.

@@

Whenever possible, we tried to parameterize our simulation model using empirical data on the spVL trait. We set the total variance in spVL to 0.73 log copies$^2$ mL$^{-2}$ based on UK cohort data \citep{Mitov2018}. Other studies have estimated slightly lower values though (Table \ref{tab:spVLmeasurements}). After allotting 25\% of this variance to a host part of spVL $h$ based on results by \citet{McLaren2015}, we partitioned the remaining variance between a viral part $g$ and an environmental part $e$ in different ratios to assess estimator performance across a range of spVL heritabilities. $h$ was simulated as the sum of contributions from 20 causal host genetic variants, 10 of which had an effect size of 0.2 log copies mL$^{-1}$ and 10 of which had an effect size of -0.2 log copies mL$^{-1}$. Host genetic variants were generated from a binomial distribution with probability $p$ calculated such that $h$ had the appropriate variance (see Table \ref{tab:sim-params}). We generated a random viral phylogeny with branch lengths on the same time scale as a previously inferred UK cohort HIV tree \citep{Hodcroft2014} using the R package ape \citep{Paradis2018}. $g$ was simulated by running an OU process along the phylogeny using the R package POUMM \citep{Mitov2017a-POUMM} and sampling values at the tips. For the OU parameters $\theta$ and $g_{0}$ we used 4.5 log copies mL$^{-1}$ based on fitting the same model to SCHS data (Table \ref{tab:POUMM-parameters-our-estimates}). This is similar to values previously inferred for HIV (Table \ref{tab:POUMMparams-prev-estimated}). To assess our estimator's performance under a range of evolutionary scenarios, we co-varied the heritability $H^2$ and selection strength $\alpha$ parameters. The intensity of random fluctuations $\sigma$ was determined based on these parameters (Table \ref{tab:sim-params}, Figure \ref{fig:sigma-vs-alpha}). Finally, the environmental part of spVL $e$ was generated from a normal distribution with mean 0.  For a full graphical model representation of the simulation scheme, see Figure \ref{fig:sim-design}.

@@

\subsection*{Phylogenetic trait correction}
We estimated the pathogen and non-pathogen effects on HIV-1 spVL and \emph{X. arboricola} mean QDR trait values using the method described in this paper. For each  individual, we estimated the pathogen part of the trait value using equation \ref{eq:MuG} and the corresponding non-pathogen part using equation \ref{eq:EHat}. This is implemented in the function ``POUMM:::gPOUMM'' in the R package POUMM. In the HIV-1 case, each sample corresponds to one HIV-1 strain with one spVL value. In the \emph{X. arboricola} case, each sample corresponds to one \emph{X. arboricola} strain and the mean QDR score for that strain across all host types (see Figure \ref{fig:application-exp-setup}). To calculate the expected correlation in trait values between tips in the pathogen phylogeny, we used the function ``covVTipsGivenTreePOUMM'' in the same package. For the POUMM parameters $\alpha$, $\sigma$, $\theta$, and $\sigma_e$, we used the posterior mean estimates generated as described above. All the code used to implement the method is available at https://github.com/cevo-public/POUMM-GWAS.

@@

>>>>

\begin{flushright} % for MBE submission
Submission type: Article\\
Section: Methods
\end{flushright}

@@

% Overview of relevant methods and why we don't use them
It is well-known that GWAS can be sensitive to confounding variables. Shared ancestry among individuals, especially between close relatives, can give rise to spurious genetic correlations with a trait. Corrections for these types of population structure in human GWAS cohorts are well-developed and widely accepted  \citep{Astle2009, Price2006}. More recently, analogous methods have been developed for microbial GWAS, where clonal reproduction exacerbates population structure \citep{Power2017}. Microbial GWAS-specific phylogenetic methods to account for population structure in microbial GWAS include explicitly testing for lineage-specific effects as in \cite{Earle2016} and modified association tests that account for phylogenetic relationships amongst samples as in \cite{collins_phylogenetic_nodate}. 
% \citep{Earle2016} simultaneously test for pathogen lineage and locus-level associations with a trait using a linear mixed model framework. \citep{collins_phylogenetic_nodate} test for locus-level associations that are significantly stronger than expected under a null model considering shared pathogen ancestry. 
These approaches are designed to quantify genetic effects from one organism on a trait.

In the infectious disease context, genetic effects from two organisms - the host and the pathogen - may affect an infectious disease trait. GWAS using paired host-pathogen genotype data have previously been done to elucidate the marginal and interaction effects of host and pathogen genetic variants. Methods to account for microbial population structure when testing for marginal host associations or host-pathogen interaction effects include adding the microbial kinship matrix as a random effect in a linear mixed model as in \cite{Wang2018Two-wayGenomes} and using principle components derived from either this matrix or the pathogen phylogeny as covariates in a linear model as in \cite{Naret2018}. These methods focus on capturing and accounting for correlations due to the pathogen phylogeny, without further investigating the nature of these correlations.
%However, the top principle components capture only some of the information from the full pathogen phylogeny. Furthermore, we would like to simultaneously address issues (i) and (ii).

@@

% Figure 1 location

@@

\begin{enumerate}
	\item Sample paired host genotypes, pathogen genome sequences, and trait values from a cohort.
	\item Construct a pathogen phylogeny using the pathogen genome sequences.
	\item Estimate the parameters of the POUMM based on the trait values and the pathogen phylogeny. This can be done with the R package POUMM \citep{Mitov2017a-POUMM}.
	\item Generate maximum-likelihood estimates for the pathogen and corresponding non-pathogen effects on the trait using equations \ref{eq:MuG} and \ref{eq:EHat}.
	\item Perform GWAS with only the non-pathogen effects on the trait as the response variable.
\end{enumerate}

@@

% Figure 2-4 location

@@

% Figure 5 location

We compared our proposed GWAS framework with a more standard approach by performing two different GWAS on the same SHCS human genotypes. We retained 1,392 individuals of European ancestry for the GWAS. In the (i) ``GWAS with standard trait value'' we used the total trait value, calculated spVL values, as the GWAS response variable. In the (ii) ``GWAS with estimated non-pathogen part of trait'' we used our estimates for the non-pathogen effects on spVL. Figure \ref{fig:gwas-results}A shows that results are qualitatively similar between the two GWAS. Q-Q plots show the distribution of p-values are very similar as well (Figure \ref{fig:qq-plots}). Figure \ref{fig:gwas-results}B shows how the strength of association changed for some variants in the MHC and \emph{CCR5} regions. Taking into account phylogenetic information slightly decreased association strength for most variants in the \emph{CCR5} region. Association strength increased for some variants in the MHC, for example, SNP rs9265880 had the greatest increase in significance in the MHC region, from a p-value of $3.5 \times 10^{-07}$ to $7.7 \times 10^{-09}$. However, the top-associated variants in the MHC and \emph{CCR5} regions were consistent regardless of the GWAS response variable used (Table \ref{tab:gwas-top-associations}). Finally, Table \ref{tab:comp-gwas-mclaren-snps} shows how our GWAS results compare for the two top-associated SNPs identified by \citet{McLaren2015}, who performed the largest standard GWAS for HIV spVL to date. Effect sizes are smaller with a phylogenetic correction and p-values are slightly increased. We repeated the analysis using three different approximate maximum-likelihood phylogenies and these results were consistent (see Materials and Methods; Table \ref{tab:comp-gwas-mclaren-snps-alt-trees}). In summary, there are no clear patterns that point to new regions of association in the human genome with spVL when we take into account the pathogen phylogeny.

% Figure 6 location

% Table 1 location

@@

Given the posterior mean estimates for the POUMM parameters, expected correlation in trait values between tips were very low (maximum value $3.2 \times 10^{-12}$ compared to maximum value of 0.45 in the HIV-1 spVL application). Thus, the phylogeny is not very informative for a trait value correction. Indeed, the estimated pathogen part of the QDR trait calculated by our method is simply a scaling of the total QDR trait value (Figure \ref{fig:trait-comparison-qdr}). We anyways selected 22 random host-pathogen strain pairings to perform a comparative GWAS analogous to that for HIV-1 spVL, where each host is infected with a single pathogen strain. In the first GWAS, we used the specific QDR measurement for each selected host-pathogen pairing. I.e., with reference to Figure \ref{fig:application-exp-setup}, we selected $z_{11}$ for the first sample, $z_{23}$ for the second sample, and so on. In the second GWAS, we used our estimates for the non-pathogen effects on QDR for each pairing. Since our method did not utilize phylogenetic information in this case, the estimated non-pathogen part of the trait is simply the specific QDR for each selected host-pathogen pairing, minus mean QDR for the respective pathogen strain, calculated across all the host \emph{A. thaliana} types. I.e., with reference to Figure \ref{fig:application-exp-setup}, we used a scaled version of $z_{11} - \bar{z}_1$ for the first sample, $z_{23} - \bar{z}_2$ for the first sample, and so on. Figure \ref{fig:gwas-results-qdr} shows that results are qualitatively similar between the two GWAS, with a slight decrease in association strength for the top-associated variants. Q-Q plots show the distribution of p-values are also very similar (Figure \ref{fig:qq-plots-qdr}). In the first, standard GWAS, one \emph{A. thaliana} loci just exceeds the threshold for significant association after correction for multiple testing. In the second, corrected GWAS, no \emph{A. thaliana} variants are significantly associated with QDR to \emph{X. arboricola}.

% Figure 7 location

@@

% Contextualize the study & address limitations
The main novelty of our approach is to estimate parameters governing the evolutionary dynamics of a trait in the pathogen population and use these estimates to correct infectious disease trait values prior to performing GWAS, thereby estimating and removing pathogen effects. In simulations, we show that when trait heritability due to shared pathogen ancestry amongst infection partners is greater than approximately 25\%, GWAS power to detect host genetic variants associated with the same trait is reduced. Our method can correct for this effect in certain evolutionary scenarios by using information from the full pathogen phylogeny. Based on our simulation results, our method is anticipated to be very useful for disease traits that are highly heritable from donor to recipient and  maintain a high correlation between sampled individuals. In simulations, we showed this is the case when pathogen heritability is high, selection strength is low, and trait values are not subject to strong stochastic fluctuations. In summary, cohort-level, phylogenetically structured differences in the measured trait value are necessary for our approach to outperform state of the art methods.

@@

Given our estimates for trait heritability and selection strength on HIV-1 spVL and \emph{A. thaliana} QDR to \emph{X. arboricola}, our simulation results reveal that we cannot expect a significant improvement in GWAS power for these systems (Figure \ref{fig:simulationResults}). Indeed, while certain pairs of samples in the HIV-1 cohort were expected to have phylogenetically correlated spvL values (maximum expected correlation between any two samples was 0.45), the overall effect on GWAS is small. For HIV-1 spVL, our phylogenetic correction slightly decreases p-values for variants in $CCR5$ and slightly decreases some and increases other p-values for variants in the MHC (Figure \ref{fig:gwas-results}B). Simulations show we shouldn't expect a net p-value decrease, but our simulations represent an ideal scenario since we simulate under the POUMM. For the empirical data, un-modeled evolutionary pressures like drug treatment and host-specific HLA alleles might cause the reduced p-values. However, the overall picture is consistent between the two GWAS (Figure \ref{fig:gwas-results}A). For \emph{A. thaliana} QDR to \emph{X. arboricola}, the trait value correction does not utilize phylogenetic information because phylogenetic correlations between samples are too weak (maximum expected correlation between strains was $3.2 \times 10^{-12}$). We anyways corrected QDR trait values based on average QDR for each pathogen strain across the full range of host types. Results show slight decrease in p-values for the most-associated variants in this application as well, but the overall picture is consistent with previous GWAS results from \cite{Wang2018Two-wayGenomes}. That study found no significant \emph{A. thaliana} variants associated with QDR using a linear mixed model jointly accounting for host genetic effects, pathogen genetic effects, and interaction effects. As with HIV-1 spVL, our results do not challenge this previous finding. Therefore, we conclude that GWAS for host determinants of HIV-1 subtype B spVL and \emph{A. thaliana} determinants of QDR to \emph{X. arboricola} are robust to our correction for pathogen effects.

Our method has several limitations. When POUMM parameter estimates are highly uncertain, correcting trait values based on posterior mean or maximum likelihood parameter estimates neglects this uncertainty. Then, as in the \emph{A. thaliana}-\emph{X. arboricola} application, fitting the POUMM may reveal that expected phylogenetic correlations between samples are not strong enough to justify using our method to correct trait values in a GWAS. In this case, one may wish to use a linear mixed model as in \cite{Wang2018Two-wayGenomes}, where the pathogen effect is co-estimated as a random effect. The expected correlation structure estimated under the POUMM could be used for the covariance of the random effect, taking the phylogeny into account differently but still utilizing information from the evolutionary model. Finally, as we show here, our method is not anticipated to be useful in certain evolutionary scenarios. For instance, traits like antimicrobial resistance may be under strong selection pressure and be highly heritable. In these instances, our simulations do not point to a large improvement when adding our pre-processing step. In any case, such traits might violate the POUMM assumption that trait values vary as a random walk in continuous space if they are caused by few mutations of strong affect, meaning our approach would not apply. In this situation, one would rather account for antimicrobial resistance as a covariate in the GWAS association model.

The primary advantage of our approach is that it is complementary to previously developed methods for infectious disease GWAS. First, it provides additional information on the evolutionary dynamics of the trait in the pathogen population. Then, it is a convenient pre-processing step for GWAS because it simply produces a corrected response variable for GWAS association tests.  In cases where a correction can be estimated and applied using our method, the corrected trait values are envisioned to be used in any of the previously developed GWAS models for the actual association testing (we used a linear model approach implemented in PLINK \citep{Chang2015}, though a more advanced method would be to use a linear mixed model with host ancestry as a random effect). Further, additional model complexity can be added to the GWAS association tests. For instance, our method does not account for co-infection, which might add additional variance to trait values and decrease GWAS power. In this case, one could add co-infection status as a covariate in the GWAS association test to account for this variable.

@@

% Summarize the paper
In summary, we present a coherent infectious disease GWAS framework that takes the pathogen phylogeny into account when searching for host determinants of a disease trait. We further show that the pathogen phylogeny only has an impact on the GWAS outputs if heritability of the trait amongst infection partners is $>$ 25\%.
For the systems studied here, spVL in individuals living with HIV and QDR for X. arboricola infections in A. thaliana, the phylogenetic correction does not change GWAS results. Our findings indicate previously published GWAS results for these systems are not biased by shared evolutionary history amongst infecting pathogen strains.

@@

Whenever possible, we tried to parameterize our simulation model using empirical data on the spVL trait. We set the total variance in spVL to 0.73 log copies$^2$ mL$^{-2}$ based on UK cohort data \citep{Mitov2018}. Other studies have estimated slightly lower values though (Table \ref{tab:spVLmeasurements}). After allotting 25\% of this variance to a host part of spVL $h$ based on results by \citet{McLaren2015}, we partitioned the remaining variance between a viral part $g$ and an environmental part $e$ in different ratios to assess estimator performance across a range of spVL heritabilities. $h$ was simulated as the sum of contributions from 20 causal host genetic variants, 10 of which had an effect size of 0.2 log copies mL$^{-1}$ and 10 of which had an effect size of -0.2 log copies mL$^{-1}$. Host genetic variants were generated from a binomial distribution with probability $p$ calculated such that $h$ had the appropriate variance (see Table \ref{tab:sim-params}). We generated a random viral phylogeny with branch lengths on the same time scale as a previously inferred UK cohort HIV tree \citep{Hodcroft2014} using the R package ape \citep{Paradis2018}. $g$ was simulated by running an OU process along the phylogeny using the R package POUMM \citep{Mitov2017a-POUMM} and sampling values at the tips. For the OU parameters $\theta$ and $g_{0}$ we used 4.5 log copies mL$^{-1}$ based on previous estimates of mean spVL (Table \ref{tab:spVLmeasurements}). This is similar to values previously inferred for HIV (Table \ref{tab:POUMMparams-prev-estimated}). To assess our estimator's performance under a range of evolutionary scenarios, we co-varied the heritability $H^2$ and selection strength $\alpha$ parameters. The intensity of random fluctuations $\sigma$ was determined based on these parameters (Table \ref{tab:sim-params}, Figure \ref{fig:sigma-vs-alpha}). Finally, the environmental part of spVL $e$ was generated from a normal distribution with mean 0.  For a full graphical model representation of the simulation scheme, see Figure \ref{fig:sim-design}.

@@

\subsection*{Phylogenetic trait correction}
We estimated the pathogen and non-pathogen effects on HIV-1 spVL in humans and \emph{X. arboricola} mean QDR in \emph{A. thaliana} using the method described in this paper. For each  individual, we estimated the pathogen part of the trait value using equation \ref{eq:MuG} and the corresponding non-pathogen part using equation \ref{eq:EHat}. This is implemented in the function ``POUMM:::gPOUMM'' in the R package POUMM. In the HIV-1 case, each sample corresponds to one HIV-1 strain with one spVL value. In the \emph{X. arboricola} case, each sample corresponds to one \emph{X. arboricola} strain and the mean QDR score for that strain across all host types (see Figure \ref{fig:application-exp-setup}). To calculate the expected correlation in trait values between tips in the pathogen phylogeny, we used the function ``covVTipsGivenTreePOUMM'' in the same package. For the POUMM parameters $\alpha$, $\sigma$, $\theta$, and $\sigma_e$, we used the posterior mean estimates generated as described above. All the code used to implement the method is available at https://github.com/cevo-public/POUMM-GWAS.

@@

% Main text table

% Main text figure legends

\begin{figure}[H]
	\begin{center}
% 	\includegraphics[width = \linewidth]{figures/simulation_setup.png}
	\caption{A high-level schematic of our phylogenetic Ornstein-Uhlenbeck mixed model (POUMM)-based simulation framework in the context of HIV-1 set-point viral load (spVL). (A) shows how the viral effects on spVL evolve along the viral phylogeny according to an Ornstein-Uhlenbeck process. (B) shows how human host genetic effects are the sum of independent effects from several causal variants. Each variant can be present in 0, 1, or 2 copies. Half the variants have a positive effect of size $\delta$ and half have a negative effect of size $\delta$. (C) shows how other environmental effects are independently drawn from a Gaussian distribution centered at 0. These three effects sum to the trait value for each simulated individual.}
		\label{fig:spVLModel}
	\end{center}
\end{figure}
